---
title: "BIOSTAT 620 Midterm Project Code"
author: "Andre Gala-Garza"
date: "3/11/2026"
format: html
editor: visual
embed-resources: true
---

**Disclaimer:** Generative AI was used to assist with templating and writing code in this assignment; however, this code was checked manually and edited by hand to ensure accuracy.

**Source:** OpenAI. (2026). *ChatGPT (GPT-5.4 Thinking)* \[Large language model\]. <https://chatgpt.com/>.

**Note:** This is NOT my midterm project; it is simply a place to store code in order to obtain results that will end up in the project report.

# Data Preparation

```{r}
library(reticulate)
py_require("anndata")
library(anndata)

ad <- read_h5ad("data.h5ad")

cols_keep <- c(
  "sample_id",
  "donor_id",
  "disease",
  "cell_type",
  "sex",
  "tissue",
  "sample_collection_site",
  "tissue_type",
  "sample_source"
)

metadata_df <- as.data.frame(ad$obs[, cols_keep])
gene_names <- as.character(ad$var_names)

genes_keep <- c(
  CXCL1 = "ENSG00000163739",
  CXCL8 = "ENSG00000169429",
  SFTPC = "ENSG00000168484",
  AGER  = "ENSG00000204305"
)

genes_present <- genes_keep[genes_keep %in% gene_names]

expr_df <- as.data.frame(ad[, unname(genes_present)]$to_df())
colnames(expr_df) <- names(genes_present)

full_df <- cbind(metadata_df, expr_df)
```

```{r}
colnames(full_df)
```

# Exploratory Analysis

## Dataset characteristics

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(stringr)

n_rows <- nrow(full_df)
n_cols <- ncol(full_df)
n_donors <- n_distinct(full_df$donor_id)
n_samples <- n_distinct(full_df$sample_id)
n_diseases <- n_distinct(full_df$disease)
n_cell_types <- n_distinct(full_df$cell_type)

data_summary <- data.frame(
  Metric = c(
    "Number of cells",
    "Number of variables",
    "Number of donors",
    "Number of samples",
    "Number of disease groups",
    "Number of cell types"
  ),
  Value = c(n_rows, n_cols, n_donors, n_samples, n_diseases, n_cell_types)
)

kable(data_summary, caption = "Dataset characteristics")
```

## Variables included

```{r}
variable_summary <- data.frame(
  Variable = colnames(full_df),
  Type = sapply(full_df, class)
)

kable(variable_summary, caption = "Variables included in the cleaned dataset")
```

## Missing data

```{r}
missing_summary <- data.frame(
  Variable = colnames(full_df),
  Missing_Count = sapply(full_df, function(x) sum(is.na(x))),
  Missing_Proportion = sapply(full_df, function(x) mean(is.na(x)))
)

kable(missing_summary, caption = "Missingness summary")
```

## Distribution of key categorical variables

```{r}
disease_counts <- full_df %>%
  count(disease, sort = TRUE)

kable(disease_counts, caption = "Cell counts by disease status")
```

```{r}
ggplot(disease_counts, aes(x = reorder(disease, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of cells by disease status",
    x = "Disease",
    y = "Cell count"
  ) +
  theme_minimal()
```

```{r}
cell_type_counts <- full_df %>%
  count(cell_type, sort = TRUE)

kable(head(cell_type_counts, 15), caption = "Top 15 cell types by count")
```

```{r}
cell_type_counts %>%
  slice_max(order_by = n, n = 15) %>%
  ggplot(aes(x = reorder(cell_type, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 15 cell types",
    x = "Cell type",
    y = "Cell count"
  ) +
  theme_minimal()
```

```{r}
sex_counts <- full_df %>%
  count(sex, sort = TRUE)

kable(sex_counts, caption = "Cell counts by sex")
```

```{r}
ggplot(sex_counts, aes(x = sex, y = n)) +
  geom_col() +
  labs(
    title = "Number of cells by sex",
    x = "Sex",
    y = "Cell count"
  ) +
  theme_minimal()
```

## Distribution of key gene expression variables

```{r}
gene_summary <- full_df %>%
  summarise(
    CXCL1_mean = mean(CXCL1, na.rm = TRUE),
    CXCL1_median = median(CXCL1, na.rm = TRUE),
    CXCL8_mean = mean(CXCL8, na.rm = TRUE),
    CXCL8_median = median(CXCL8, na.rm = TRUE),
    SFTPC_mean = mean(SFTPC, na.rm = TRUE),
    SFTPC_median = median(SFTPC, na.rm = TRUE),
    AGER_mean = mean(AGER, na.rm = TRUE),
    AGER_median = median(AGER, na.rm = TRUE)
  )

kable(gene_summary, caption = "Summary statistics for selected genes")
```

```{r}
ggplot(full_df, aes(x = CXCL1)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of CXCL1 expression") +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = CXCL8)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of CXCL8 expression") +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = SFTPC)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of SFTPC expression") +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = AGER)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of AGER expression") +
  theme_minimal()
```

# Preliminary Results

## Cell type composition across disease groups

```{r}
disease_celltype_counts <- full_df %>%
  count(disease, cell_type) %>%
  group_by(disease) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

kable(
  head(disease_celltype_counts, 20),
  caption = "Cell type counts and proportions by disease"
)
```

```{r}
disease_celltype_counts %>%
  group_by(cell_type) %>%
  summarise(total = sum(n)) %>%
  slice_max(order_by = total, n = 10) %>%
  pull(cell_type) -> top_cell_types

full_df %>%
  filter(cell_type %in% top_cell_types) %>%
  count(disease, cell_type) %>%
  group_by(disease) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = disease, y = prop, fill = cell_type)) +
  geom_col() +
  labs(
    title = "Cell type composition by disease status",
    x = "Disease",
    y = "Proportion"
  ) +
  theme_minimal()
```

## Gene expression by disease

```{r}
gene_disease_summary <- full_df %>%
  group_by(disease) %>%
  summarise(
    n = n(),
    CXCL1_mean = mean(CXCL1, na.rm = TRUE),
    CXCL1_median = median(CXCL1, na.rm = TRUE),
    CXCL8_mean = mean(CXCL8, na.rm = TRUE),
    CXCL8_median = median(CXCL8, na.rm = TRUE),
    SFTPC_mean = mean(SFTPC, na.rm = TRUE),
    SFTPC_median = median(SFTPC, na.rm = TRUE),
    AGER_mean = mean(AGER, na.rm = TRUE),
    AGER_median = median(AGER, na.rm = TRUE)
  )

kable(gene_disease_summary, caption = "Summary statistics for selected genes by disease status")
```

```{r}
ggplot(full_df, aes(x = disease, y = CXCL1)) +
  geom_boxplot() +
  labs(
    title = "CXCL1 expression by disease status",
    x = "Disease",
    y = "CXCL1 expression"
  ) +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = disease, y = CXCL8)) +
  geom_boxplot() +
  labs(
    title = "CXCL8 expression by disease status",
    x = "Disease",
    y = "CXCL8 expression"
  ) +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = disease, y = SFTPC)) +
  geom_boxplot() +
  labs(
    title = "SFTPC expression by disease status",
    x = "Disease",
    y = "SFTPC expression"
  ) +
  theme_minimal()
```

```{r}
ggplot(full_df, aes(x = disease, y = AGER)) +
  geom_boxplot() +
  labs(
    title = "AGER expression by disease status",
    x = "Disease",
    y = "AGER expression"
  ) +
  theme_minimal()
```

## Follow-up analysis: gene expression by disease within cell type

```{r}
sort(unique(full_df$cell_type))
```

```{r}
full_df %>%
  filter(cell_type == "pulmonary alveolar type 2 cell") %>%
  group_by(disease) %>%
  summarise(
    n = n(),
    SFTPC_mean = mean(SFTPC, na.rm = TRUE),
    SFTPC_median = median(SFTPC, na.rm = TRUE)
  ) %>%
  kable(caption = "SFTPC summary statistics in pulmonary alveolar type 2 cells by disease")
```

```{r}
full_df %>%
  filter(cell_type == "pulmonary alveolar type 2 cell") %>%
  ggplot(aes(x = disease, y = SFTPC)) +
  geom_boxplot() +
  labs(
    title = "SFTPC expression in alveolar type 2 cells by disease",
    x = "Disease",
    y = "SFTPC expression"
  ) +
  theme_minimal()
```

# Conclusion

```{r}
```
